The imprint of global magma oceans on exoplanet demographics

### 2023-03-17 src/tex/ms.tex
test different M-R relations, add plot and discuss differences

\subsection{Orbit parameters and planet masses}
Eccentric orbits alter the probability of a planet to transit \citep[e.g.,][]{Barnes2007a}.
The distribution of eccentricities $e$ of exoplanets has been found to resemble a Beta function \citep{Kipping2013b}, which we chose to draw synthetic eccentricities from.
Following~\citet{Kipping2013b}, we used a Beta distribution with parameters $a=0.867$ and $b=3.03$, and truncated the distribution at $e = 0.8$.
%Another parameter with an even higher impact on transit probabilities is the orbital inclination toward the observer $i$.
Assuming isotropic alignments of orbits, we assigned each planet an inclination drawn from a distribution uniform in $\cos(i)$.

@@

\section{Results}
\subsection{Statistical signature of the runaway greenhouse threshold}\label{sec:res_signature}
To characterize the population-level imprint of individual radius changes, we generated a generic planet population with an injected runaway greenhouse effect assuming a water fraction $x_\mathrm{H_2O} = \var{wrr}$.
\begin{figure}
    \begin{centering}
        \includegraphics[width=\hsize]{figures/HnullHmo.pdf}
        \caption{Synthetic planets above and below the runaway greenhouse threshold.
        \textit{Top}: Planet state as a function of orbital period. Planets with and without a runaway greenhouse climate mix and are not distinguishable in period space.
        \textit{Bottom}: Radii of synthetic planets with injected radius deviation as a function of net instellation. Only the planets marked as transiting are observable.
            Above the runaway greenhouse threshold $S_\mathrm{thresh} =\SI{280}{\watt\per\meter\squared} $, some planets maintain their original radii (green crosses) whereas some have their radii inflated (blue crosses) by the amount indicated with gray lines.
        The sharp boundary at $S_\mathrm{thresh}$ causes a discontinuity in the average planet radius (blue line).
            This runaway greenhouse hypothesis can be tested against the null hypothesis $H_0$ (green line), where average radii are independent of instellation.}
        \label{fig:HnullHmo}
    \end{centering}
\end{figure}
Figure~\ref{fig:HnullHmo} shows the resulting planetary radii.
When ordered in orbital period space, the different planet types overlap, diluting the demographic imprint.
With net instellation as an independent variable, planets above and below the runaway greenhouse threshold separate: The runaway greenhouse-induced radius inflation introduces a discontinuity of average planet radii and bulk densities as a function of stellar irradiation.
We also show the predictions of observable average planet radii from the statistical hypotheses defined above.
%The value for the average radius \textit{outside} the runaway greenhouse regime was computed from our nominal observed sample, whereas the radius change in the magma ocean case compared to the null hypothesis is a prediction from the magma ocean model, evaluated on the same sample.
Within the runaway greenhouse regime, an average radius change of \avgRadiusChange\ occurs.

@@

\subsubsection{Baseline mass-radius relationship}
Our baseline mass-radius relationship assuming pure $\mathrm{MgSiO_3}$ interiors~\citep{Zeng2016} might not be representative of the rocky planet population.
However, while interior composition may change the magnitude of radius inflation and could introduce an offset to its demographic imprint, we do not expect a change of its structure.
The magnitude of the radius inflation effect is in fact expected to be larger for an Earth-like interior composition with an iron core-silicate mantle structure.\todo[inline]{TIM: do you have a reference for this statement?}
We thus consider our mass-radius relationship a conservative case and do not expect a negative impact on the detectability of the runaway greenhouse transition due to deviations from it.
\todo[inline]{if feasible: run a test with a different M-R relation. do we expect that the choice of M-R relation impacts the results? Perhaps have an appendix section.}

@@

\subsubsection{``Sharpness'' of the demographic trend}
The real counterpart to our idealized step-function central to our hypothesis tests could be blurred by several factors.
Its amplitude is affected by ``sub-Neptunes'' with significant Hydrogen/Helium envelopes, which can disguise as inflated rocky planets.
They might not contribute to the demographic signal we are looking for, although a similar transition was suggested for water-dominated sub-Neptunes~\citep{Pierrehumbert2022}.
A subset of such gas-rich planets will experience atmospheric loss via photoevaporation~\citep{Owen2013} or core-powered mass loss~\citep{Ginzburg2018}, turning them from planets appearing inflated into much smaller \mbox{(super-)Earths}.
Further dilution comes from the population of planets that lack an atmosphere or sufficient volatiles that can create a steam atmosphere.
There is also a time dependence, both because stellar luminosities change with age and because runaway greenhouse planets with limited water reservoirs ultimately desiccate, ending their steam atmosphere phase.
Intrinsic variation in the threshold instellation is caused by differences in planetary features influencing the onset of a runaway climate such as albedo, atmospheric composition, or surface gravity~\citep{Pierrehumbert2022}.

@@

>>>>

\subsection{Orbit parameters and planet masses}\label{sec:met-orbits_masses}
Eccentric orbits alter the probability of a planet to transit \citep[e.g.,][]{Barnes2007a}.
The distribution of eccentricities $e$ of exoplanets has been found to resemble a Beta function \citep{Kipping2013b}, which we chose to draw synthetic eccentricities from.
Following~\citet{Kipping2013b}, we used a Beta distribution with parameters $a=0.867$ and $b=3.03$, and truncated the distribution at $e = 0.8$.
%Another parameter with an even higher impact on transit probabilities is the orbital inclination toward the observer $i$.
Assuming isotropic alignments of orbits, we assigned each planet an inclination drawn from a distribution uniform in $\cos(i)$.

@@

\section{Results}
\subsection{Statistical signature of the runaway greenhouse threshold}\label{sec:res_signature}
To characterize the population-level imprint of individual radius changes, we generated a generic planet population with an injected runaway greenhouse effect assuming a water fraction $x_\mathrm{H_2O} = \var{wrr}$.
\begin{figure}
    \begin{centering}
        \includegraphics[width=\hsize]{figures/HnullHmo.pdf}
        \caption{Synthetic planets above and below the runaway greenhouse threshold.
        \textit{Top}: Planet state as a function of orbital period. Planets with and without a runaway greenhouse climate mix and are not distinguishable in period space.
        \textit{Bottom}: Radii of synthetic planets with injected radius deviation as a function of net instellation. Only the planets marked as transiting are observable.
            Above the runaway greenhouse threshold $S_\mathrm{thresh} = \var{injectedSthresh}$, some planets maintain their original radii (green crosses) whereas some have their radii inflated (blue crosses) by the amount indicated with gray lines.
        The sharp boundary at $S_\mathrm{thresh}$ causes a discontinuity in the average planet radius (blue line).
            This runaway greenhouse hypothesis can be tested against the null hypothesis $H_0$ (green line), where average radii are independent of instellation.}
        \label{fig:HnullHmo}
    \end{centering}
\end{figure}
Figure~\ref{fig:HnullHmo} shows the resulting planetary radii.
When ordered in orbital period space, the different planet types overlap, diluting the demographic imprint.
With net instellation as an independent variable, planets above and below the runaway greenhouse threshold separate: The runaway greenhouse-induced radius inflation introduces a discontinuity of average planet radii and bulk densities as a function of stellar irradiation.
We also show the predictions of observable average planet radii from the statistical hypotheses defined above.
%The value for the average radius \textit{outside} the runaway greenhouse regime was computed from our nominal observed sample, whereas the radius change in the magma ocean case compared to the null hypothesis is a prediction from the magma ocean model, evaluated on the same sample.
Within the runaway greenhouse regime, an average radius change of \avgRadiusChange\ occurs.

@@

\subsubsection{Baseline mass-radius relationship}
Our baseline mass-radius relationship assuming pure $\mathrm{MgSiO_3}$ interiors~\citep{Zeng2016} might not be representative of the rocky planet population.
However, while interior composition may change the magnitude of radius inflation and could introduce an offset to its demographic imprint, we do not expect a change of its structure.
The magnitude of the radius inflation effect is in fact expected to be larger for an Earth-like interior composition with an iron core-silicate mantle structure.\todo[inline]{TIM: do you have a reference for this statement?}
We thus consider our mass-radius relationship a conservative case and do not expect a negative impact on the detectability of the runaway greenhouse transition due to deviations from it.
Nevertheless, we performed a sanity check to assess the influence varying our baseline model on our results (see Appendix~\ref{app:MR_relation}), and found general agreement between different interior composition models.

@@

\subsubsection{``Sharpness'' of the demographic trend}
The real counterpart to our idealized step-function central to our hypothesis tests could be blurred by several factors.
Its amplitude is affected by ``sub-Neptunes'' with significant Hydrogen/Helium envelopes, which can disguise as inflated rocky planets.
They might not contribute to the demographic signal we are looking for, although a similar transition was suggested for water-dominated sub-Neptunes~\citep{Pierrehumbert2022}.
A subset of such gas-rich planets will experience atmospheric loss via photoevaporation~\citep{Owen2013} or core-powered mass loss~\citep{Ginzburg2018}, turning them from planets appearing inflated into much smaller \mbox{(super-)Earths}.
Further dilution comes from the population of planets that lack an atmosphere or sufficient volatiles that can create a steam atmosphere.
There is also a time dependence, both because stellar luminosities change with age and because runaway greenhouse planets with limited water reservoirs ultimately desiccate, ending their steam atmosphere phase.

Intrinsic variation in the threshold instellation is caused by differences in planetary features influencing the onset of a runaway climate such as albedo, atmospheric composition, or surface gravity~\citep{Pierrehumbert2022}.
The choice of statistical estimator for the hypothesis tests may further blur the recovered discontinuity.
We compare our nominal running mean approach with a binned statistic in Appendix~\ref{app:binnedstats}.

@@

\subsection{Influence of different mass-radius relationships on our results}\ref{app:MR_relation}
We assessed how a different choice of baseline mass-radius relation influences our results.
Focussing on the detectability of the statistical runaway greenhouse signal and the ability to constrain the threshold instellation, we repeated the hypothesis test in Sect.~\ref{sec:res_testability} with alternative mass-radius relationships.
Instead of assuming a pure $\mathrm{MgSiO_3}$ composition, we assigned planet masses using either the probabilistic relationship in \citet{Wolfgang2016} or a semi-empirical, two-layer relation assuming an Earth-like (\SI{32.5}{\percent} Fe + \SI{67.5}{\percent} MgSiO$_3$) composition~\citep{Zeng2016}.

\begin{figure}[ht!]
%    \script{figurescript.py}
    \begin{centering}
        \includegraphics[width=\linewidth]{figures/MR-violins.pdf}
        \caption{
        Posterior threshold instellations as a function of dilution factor for two alternative baseline mass-radius relations.
        For each grid step in $f_\mathrm{rgh}$, we show kernel density estimates of retrieved posteriors (averaged over 50 iterations) assuming an Earth-like (\SI{32.5}{\percent} Fe + \SI{67.5}{\percent} MgSiO$_3$) composition or the probabilistic relation from \citet{Wolfgang2016} and otherwise following the optimistic scenario in Sect.~\ref{sec:res_testability}.
        Lines within the violins show quartiles of the distributions, and the gray line indicates the injected threshold instellation of \var{injectedSthresh}.
        Higher log-evidence differences $\Delta \ln Z$ correspond to more significant rejections of the null hypothesis.
        In the regime of strong detections, both mass-radius relations lead to similar, accurate constraints on $S_\mathrm{thresh}$.
        Differences occur at low dilution factors, where the Earth-like relation leads to narrower estimates.
        }
        \label{fig:MR-violins}
    \end{centering}
\end{figure}
Figure~\ref{fig:MR-violins} shows how the two alternative baseline mass-radius relations influence the significance of a detection and the ability to constrain the threshold instellation.
The relation of \citet{Wolfgang2016} includes intrinsic scatter, which impedes a detection at low dilution factors.
In this regime, a layered, Earth-like composition leads to more significant detections and a narrower, although biased, constraint on $S_\mathrm{thresh}$.
Both mass-radius relations agree and recover the injected value where the null hypothesis can be rejected with high significance.
This is consistent with our nominal mass-radius relation (compare Sect.~\ref{sec:met-orbits_masses}).
We conclude that the underlying core and mantle composition of planets may affect the detectability of the transition if the fraction of planets with runaway greenhouse climates is low; however, the overall trends that our experiments revealed appear robust.

We caution that this analysis may serve only as a sanity check and should not be taken as a result in itself: The atmospheric model from \citet{Turbet2020} we adopted relies on a silicate interior composition for its transit radius prediction.
Therefore, only our nominal procedure throughout the main body of the paper represents a self-consistent treatment.
